\section{Implementation}
\label{sec:implementation}

The MI engine is implemented in Rust ($\sim$10,000 lines, 180+ unit tests). The choice of Rust is motivated by three requirements: (i) memory safety without GC pauses, (ii) zero-cost abstractions for the inner simulation loop, and (iii) safe concurrency for parallel morphon updates. The crate compiles to native, WebAssembly (for an interactive in-browser visualizer), and Python bindings via PyO3 (for benchmark harnessing).

\subsection{Core Data Structures}

Morphons are stored in a \texttt{HashMap} keyed by stable IDs. The topology is a directed graph using the \texttt{petgraph} crate, with synapses as edge weights. We use generational indices so that references survive birth/death events.

\subsection{Pulse Kernel Lite: Cache-Resident Fast Path}
\label{sec:pulse_kernel}

The inner simulation loop (spike integration, threshold, refractory) must execute every tick for all active morphons. The naive array-of-structs (AoS) layout - one \texttt{Morphon} struct per unit, $\sim$480 bytes each - places the hot fields (voltage, threshold, fired, refractory) far apart in memory. At 300 morphons this scatters $\sim$60~KB across cache lines, causing frequent L1 misses on every tick.

\emph{Pulse Kernel Lite} separates the fast-path fields into seven dense \texttt{Vec<f32>}/\texttt{Vec<bool>} arrays that sit alongside the \texttt{Morphon} structs:

\begin{center}
\texttt{voltage} | \texttt{threshold} | \texttt{input\_current} | \texttt{fired} | \texttt{refractory} | \texttt{leak\_decay} | \texttt{energy}
\end{center}

At 300 morphons, all seven arrays fit in $\sim$9.6~KB - within a typical 32~KB L1 data cache. The source of truth remains the \texttt{Morphon} structs; a lightweight sync protocol propagates changes bidirectionally:
\begin{itemize}
  \item \textbf{Fast tick}: hot arrays only. Input currents extracted, voltage integrated, threshold checked, energy gated.
  \item \textbf{Medium tick (every 10)}: \texttt{sync\_hot\_to\_structs()} before metabolic updates;\newline \texttt{sync\_structs\_to\_hot()} after.
  \item \textbf{Slow/glacial ticks}: \texttt{rebuild\_hot\_arrays()} after structural changes (synaptogenesis, division, apoptosis).
\end{itemize}

On the MNIST v2 workload (3000 images, 3 epochs, $\sim$1260 morphons), this reduces wall time from 568~s (V2, GlobalKWTA) to 130~s (V3, LocalInhibition) - a \textbf{4.4$\times$ speedup} - with no change in accuracy. The speedup compounds with network size because the fraction of time spent in the inner loop grows with $N$.

\subsection{The Main Loop}

\begin{algorithm}
\caption{System::step (one fast tick)}
\label{alg:step}
\begin{algorithmic}[1]
\State \textbf{1.} Spike propagation: deliver pending spikes from previous step
\State \textbf{2.} Inhibitory competition (k-WTA or local interneurons)
\State \textbf{3.} Morphon update: integrate inputs, decide firing, pay metabolic cost
\State \textbf{4.} Generate new spikes from fired morphons
\If{tick.medium}
  \State \textbf{5.} Update eligibility traces, apply three-factor learning
  \State \textbf{6.} Run Endoquilibrium tick (sense vitals, adjust channels)
\EndIf
\If{tick.slow}
  \State \textbf{7.} Synaptogenesis, pruning, migration
\EndIf
\If{tick.glacial}
  \State \textbf{8.} Division, differentiation, fusion, apoptosis
\EndIf
\If{tick.homeostasis}
  \State \textbf{9.} Synaptic scaling, inter-cluster inhibition
\EndIf
\end{algorithmic}
\end{algorithm}

\subsection{Output Pathways}
\label{sec:output_pathways}

MI provides two output pathways. The spike-based pathway reads motor morphon potentials after propagation and drives action selection in control tasks (CartPole). The analog readout bypasses spike conversion and reads associative morphon potentials directly: $V_j = b_j + \sum_i w_{ji} \cdot (\sigma(p_i) - 0.5)$, biologically motivated by Purkinje cell analog dendritic integration. The readout engineering and its experimental validation are described in Section~\ref{sec:readout_engineering} alongside the CartPole results where it was developed.

\subsection{Inhibitory Spike-Timing-Dependent Plasticity}
\label{sec:istdp}

Local inhibition (Section~\ref{sec:exp_mnist}) requires the inhibitory interneuron weights to self-organize without explicit k-WTA supervision. We implement the symmetric iSTDP rule of \citet{vogels2011inhibitory}, applied per-delivered-spike rather than per-medium-tick:

\paragraph{Pre-spike update.} When an inhibitory spike arrives at excitatory neuron $j$:
\begin{equation}
  \Delta w_{ij} = -\eta \cdot x_j^{\text{post}}
\end{equation}
where $x_j^{\text{post}}$ is the post-synaptic trace. An active excitatory neuron ($x_j > 0$) that receives an inhibitory spike immediately strengthens that inhibitory connection.

\paragraph{Post-spike update.} When excitatory neuron $j$ fires:
\begin{equation}
  \Delta w_{ij} = -\eta \cdot (x_i^{\text{pre}} - \alpha), \quad \alpha = 2\rho_0\tau
\end{equation}
where $x_i^{\text{pre}}$ is the pre-synaptic trace of the inhibitory interneuron and $\rho_0$ is the target firing rate. When $x_i^{\text{pre}} > \alpha$ (inhibitory interneuron has been more active than the target rate warrants), inhibition strengthens; when $x_i^{\text{pre}} < \alpha$, it weakens. The $\alpha$ term sets the equilibrium point without requiring any global coordination.

Weights are clamped to $[-5.0,\,-0.001]$ - strictly negative, enforcing Dale's law for inhibitory connections.

The critical difference from the rate-based predecessor is \emph{temporal resolution}. Rate-based iSTDP (fired once per 10 medium ticks, using a 100-tick activity average) could not react to within-burst hub activity; spike-timing iSTDP responds on the timescale of individual spike arrivals, matching the biological timescale of inhibitory short-term plasticity \citep{vogels2011inhibitory}.

\subsection{Three-Factor Learning Rule}

The three-factor STDP rule on the medium path:
\begin{equation}
\Delta w_{ij} = \eta_{ij} \cdot e_{ij}(t) \cdot M_j(t) \cdot g_a
\end{equation}
where $e_{ij}$ is the eligibility trace, $M_j$ is the receptor-gated modulator signal at the post-synaptic morphon, $\eta_{ij}$ is a per-synapse plasticity rate, and $g_a$ is an astrocytic gate.

The eligibility trace integrates STDP coincidences:
\begin{equation}
\frac{de_{ij}}{dt} = -\frac{e_{ij}}{\tau_e} + A_+ \cdot \text{tr}_i^{\text{pre}} \cdot \delta_j^{\text{post}} - A_- \cdot \text{tr}_j^{\text{post}} \cdot \delta_i^{\text{pre}}
\end{equation}
with weight-dependent bounds \cite{gilson2011stability} that make weak synapses easy to strengthen and strong synapses easy to weaken.

\subsection{Tag and Capture}

We implement the synaptic tagging-and-capture model of Frey and Morris \cite{frey1997synaptic} with one critical departure: capture is \emph{episode-gated}, not per-tick. The naive per-tick rule (``capture when reward $>$ threshold'') fails on RL tasks because the reward signal saturates at 1.0 during \emph{any} successful CartPole step - every alive step captures everything, the weight distribution collapses to uniformity (weight std drops from 1.5 to 0.15), and learning stops. The fix is to defer the capture decision to episode end and gate it on relative performance: if this episode beat the running average, capture the tagged synapses proportional to how much it beat the average; if it underperformed, decay the tags instead.

Concretely: synapses with high recent eligibility set a tag; tags accumulate at a fixed local rate (\texttt{tag\_accumulation\_rate}=0.3) and decay slowly ($\tau_{\text{tag}}=200$ steps). The \texttt{consolidation\_level} field on each synapse is a continuous $[0, 1]$ value (replacing the binary ``consolidated'' flag), and at episode end, tagged synapses gain consolidation proportional to performance delta. Consolidated synapses retain residual plasticity (10\% at \texttt{consolidation\_level}=1.0) so the network can still adapt without destroying learned features.

Endoquilibrium gates the global capture rate via a dedicated \texttt{consolidation\_gain} channel. This separation matches the biology: \emph{tags are local and continuous} (driven by per-synapse STDP eligibility), \emph{capture is global and episodic} (driven by neuromodulatory protein-synthesis-related signals during salient events). The earlier design that combined them under a single ``endo\_tag\_factor'' was biologically incorrect and produced an unstable feedback loop.

\subsection{ANCS-Core: Native Adaptive Memory Substrate}
\label{sec:ancs}

The Adaptive Node-Cluster System (ANCS) is implemented natively in Rust (\texttt{src/ancs.rs}) as the long-term memory substrate. It replaces an earlier architecture that forwarded memory operations to a separate HTTP/PostgreSQL service; ANCS runs entirely in-process with zero serialization overhead.

\paragraph{Architecture.}
The core abstraction is a \texttt{MemoryBackend} trait implemented by \texttt{InMemoryBackend}, which maintains a flat vector of \texttt{MemoryItem}s indexed by importance score. Items are organized into four \emph{VBC-lite tiers}: Verbatim (exact episode records), Structural (connectivity snapshots), Semantic (abstracted associations), and Procedural (morphon-count timelines). Item importance is scored by six factors following the SM-2 spaced-repetition model \cite{wozniak1990sm2}: centrality, novelty, reward correlation, consolidation level, recency, and access frequency.

\paragraph{Epistemic lifecycle.}
Each memory item carries an epistemic state through a six-stage lifecycle: Hypothesis $\to$ Supported $\to$ Stale $\to$ (Suspect $|$ Contested) $\to$ Outdated. Conflict detection (TruthKeeper) flags items as Contested when a new item shares Jaccard similarity $\geq 0.3$ with an existing item but carries an opposing reward sign. Conflicts trigger a cascade that re-evaluates dependent items (blast-radius propagation), maintaining a self-consistent episodic record.

\paragraph{Integration with System.}
\texttt{System} carries an \texttt{InMemoryBackend} and a \texttt{SystemHeartbeat} (a cache-line-aligned atomic counter at \texttt{\#[repr(C, align(64))]}). On each memory tick, ANCS advances, TruthKeeper reconsolidates disputed items, and the heartbeat updates. Episodic encoding creates ANCS items from the current firing pattern; \texttt{SynapticJustification.memory\_item\_ref} links individual synapses to the ANCS item that last updated them, enabling targeted weight-level episodic replay. \texttt{System::retrieve\_memory()} fuses TripleMemory and ANCS results via Reciprocal Rank Fusion (RRF); \texttt{dream\_cycle()} replays the top-5 highest-importance ANCS items to re-consolidate learned associations during quiescent periods.

\subsection{Reproducibility}

The full benchmark suite is reproducible from a clean checkout:
\begin{verbatim}
cargo run --example cartpole --release
cargo run --example mnist_v2 --release
\end{verbatim}
Each example saves a JSON record of its configuration and results to \texttt{docs/\allowbreak benchmark\_results/\allowbreak v\{version\}/}. All numbers reported in this paper are from these JSON records.
