\section{Discussion}
\label{sec:discussion}

\subsection{The Spike-vs-Analog Gap}

The most striking architectural finding is the gap between spike-based and analog readout, documented in Section~\ref{sec:exp_mnist}. The same trained network produces chance accuracy via a spike-based motor readout and high accuracy via an analog readout over the same morphon potentials. The information is present in the potentials --- the analog readout proves it --- but the spike pipeline destroys it through propagation delays, leaky integration, and multi-hop conversion.

This is not a bug in MI; it is a property of any spiking architecture in which the readout pathway sums potentials through several intermediate firing decisions. Biology solves this by having dedicated analog readout pathways for tasks that require it (e.g.\ Purkinje cells in the cerebellum perform analog integration of climbing fiber and parallel fiber inputs to produce a continuous motor signal; \cite{ito2006cerebellar}). MI's analog readout is the in-silico analogue.

The spike pipeline still has a job: it produces the diverse, sparse, energy-efficient hidden representations that the analog readout exploits. The two pathways are complementary, not competing. The MI thesis is not ``replace spikes with analog'' but ``spike-based developmental dynamics produce features that analog readouts can use.''

\subsection{Identifying and Resolving LSM Co-adaptation}

Initial post-training diagnostics (v4.4.0) revealed an apparent paradox: the system achieved 48.5\% online while offline classifiers on the same frozen activations scored 10.5\%. The sensory-only ablation quantified this: zeroing associative readout weights reduced accuracy by 35.5pp, confirming the reservoir was doing real work --- but only in conjunction with the co-trained readout weights. The system was operating in the Liquid State Machine (LSM) regime \cite{maass2002real,lukosevicius2009reservoir}: features were discriminative only through the co-adapted readout context, not as transferable representations.

The accuracy ceiling at $\sim$50\% was accordingly a co-adaptation boundary. Notably, reservoir scaling and data scaling both failed to move it (Section~\ref{sec:exp_mnist}): adding more morphons and training on up to 10K images each produced no improvement or mild regression. The bottleneck was not capacity or data --- it was the representational regime itself.

\paragraph{Diagnosing the mechanism.}
A sequential-versus-shuffled evaluation diagnostic confirmed the mechanism precisely. Evaluating V3 on a randomised test-set order (state carrying over between images) degraded accuracy substantially relative to the natural sequential order, while stateless evaluation (state reset before each image) yielded \textbf{44\%} versus \textbf{18\%} online. The system was accumulating transient state across images that biased the readout toward recent sequential context rather than per-image discriminative features. The readout was, in effect, co-adapted to sequential state as much as to actual image content.

\paragraph{Stateless training resolves co-adaptation.}
The fix is architecturally minimal: reset all transient network state (membrane potentials, working memory, firing counts) before each training image. The network then cannot exploit inter-image temporal context, and the readout must learn per-image discriminative representations. We call this \emph{stateless training} (V3-SL).

The effect is large. On the fast benchmark profile (500 training images, 2 epochs), V3-SL achieves \textbf{77\%} stateless accuracy versus \textbf{44\%} for standard V3 --- a 33pp gain on the same architecture, same training data, same number of parameters. On the standard profile (5{,}000 training images, 3 epochs), V3-SL achieves \textbf{87.7\%}, exceeding the roadmap target of 70\% and establishing a new accuracy baseline for the MI engine. Moreover, the gap between online accuracy and stateless accuracy narrows substantially under stateless training, confirming that the remaining online deficit is residual sequential state carry-over during evaluation, not a fundamental incapacity.

\paragraph{Interaction with the novelty gain fix.}
A concurrent engineering fix addressed a second bottleneck: the Endoquilibrium Rule~7 (energy pressure) was suppressing $\mathit{ng}$ (novelty gain) to $\sim$0.32 during training. Morphon energy depletes under continuous firing; the original rule responded by multiplying $\mathit{ng}$ by 0.2 under emergency energy conditions, throttling the three-factor STDP signal mid-training. The fix preserves $\mathit{ng}$ under energy pressure (only \texttt{plasticity\_mult} is suppressed) and holds $\mathit{ng}$ at $\sim$1.60 throughout training on both MNIST and CartPole. The combined effect of stateless training and stable $\mathit{ng}$ is what yields the 77\% fast-profile result; the two fixes are synergistic rather than redundant.

This reframes the Spike-vs-Analog discussion above. The analog readout's advantage is not only that it avoids lossy spike propagation; it is that it is the co-adapted partner of the reservoir the system learned under stateless training. With sequential training, the readout had two targets simultaneously: per-image classification and sequential temporal context. Stateless training removes the second target, concentrating the readout on the first.

\subsection{Self-Healing as a Diagnostic}

The v3.0.0 codebase showed what appeared to be beneficial self-healing: a 30\% intact network recovered to 48\% after 30\% ablation, an 18pp gain over its pre-damage state. This result is real and reproducible in that codebase. In v4.1.0 it disappears: the intact network reaches 52\% and post-damage recovery falls to 28.5\%.

The mechanism is now clear. In v3.0.0, rate-based iSTDP failed to suppress hub morphons during training. Hubs entrenched while Endoquilibrium's Mature stage gate froze plasticity, leaving the system stuck at 30\%. Damage removed hubs at random; Endoquilibrium re-entered the Differentiating stage in response to the performance drop, providing the high-plasticity window the original training never sustained; regrowth filled the gap with well-specialized morphons. The +18pp came entirely from two sources: hub removal and plasticity restoration --- exactly what spike-timing iSTDP provides automatically.

This reframes the v3.0.0 self-healing result as a diagnostic rather than a feature. It told us: (1) hubs are the bottleneck; (2) the substrate is capable of 48\% if given the right plasticity regime. Spike-timing iSTDP is the engineered solution to both. The v4.1.0 intact ceiling (52\%) is not coincidentally above the v3.0.0 post-recovery ceiling (48\%); it is the same substrate, finally given the conditions it needed.

The general principle is: \emph{when a developmental system benefits from damage, the damage is revealing a regulatory failure, not demonstrating a useful property}. The right response is to find and fix the failure, not to incorporate damage into the training protocol.

\subsection{Limitations}

\begin{itemize}
  \item \textbf{Online accuracy remains below state of the art.} The stateless accuracy on the standard profile (87.7\%) is competitive with biologically-plausible unsupervised STDP approaches \cite{diehl2015unsupervised} but below supervised SNNs trained with surrogate gradients, which achieve $>$98\%. The online accuracy (sequential evaluation) is lower still, because residual inter-image state carry-over during evaluation introduces noise. A full state-reset evaluation protocol --- which we now argue is the correct measure for a per-image classifier that happens to be implemented on a stateful substrate --- removes this confound and yields the stateless figure.
  \item \textbf{No language tasks yet.} The system can encode and retain information across sequential inputs (temporal memory exists in residual potentials), but compositional reasoning is not yet demonstrated. Composition requires nonlinear hidden features formed under DFA-driven supervised feedback, which the current spike pipeline does not deliver.
  \item \textbf{The premature Mature fix is a patch, not a principled solution.} The history gate at 2000 ticks prevents the failure but is essentially ``disable Mature stage detection.'' The principled fix is RPE-based regulation via a Limbic Circuit module that does not yet exist.
  \item \textbf{Performance.} The v4.1.0 implementation processes $\sim$23 images/second with $\sim$74K synapses (130~s for 3000 images), a 4.4$\times$ improvement over v3.0.0 via Pulse Kernel Lite. The medium path (eligibility traces, weight updates) is now the dominant cost; sparse eligibility updates and parallel medium-path learning are the next target.
\end{itemize}

\subsection{The Path to Language}

The planned trajectory beyond this paper:
\begin{enumerate}
  \item \textbf{Local inhibitory competition with iSTDP \cite{vogels2011inhibitory}.} \textit{Implemented in v4.1.0.} Hub dominance broken (1 labeled class $\rightarrow$ 12, accuracy 8\% $\rightarrow$ 44.5\%). Spike-timing iSTDP per delivered spike; 4.4$\times$ wall-clock speedup via Pulse Kernel Lite hot arrays. \textit{v4.6.0:} Stateless training resolves LSM co-adaptation (44\% $\rightarrow$ 77\% stateless on fast profile; 87.7\% on standard). Novelty gain stabilization ($\mathit{ng}=1.60$, was collapsing to 0.32) synergizes with stateless training.
  \item \textbf{Limbic Circuit + RPE-based regulation.} Replace the history-gate patch with reward prediction error feeding a Motivational Drive module. This naturally handles both RL and classification.
  \item \textbf{Temporal sequence processing.} Recurrent connectivity within associative populations + working-memory feedback into sensory ports. Spec exists; implementation gated on MNIST $\geq$ 85\%.
  \item \textbf{Tokenization and embedding.} Once the temporal pipeline works, add a learned tokenizer and vocabulary embedding. This is the bridge to actual NLP tasks.
\end{enumerate}

The structural decisions for each step are documented in the project's specification corpus. The question this paper does \emph{not} attempt to answer is whether MI will ever match Transformer performance on language tasks. We argue only that developmental dynamics are a principled and underexplored axis of neural architecture design, and that the failure modes we encountered have biological analogues whose biology-derived fixes work.
