\section{Conclusion}
\label{sec:conclusion}

We have presented Morphogenic Intelligence, a neural architecture that treats network topology as a learned artifact rather than a fixed design choice. Morphons are born, differentiate, fuse, and die under local rules grounded in six biological principles. The full system is implemented in Rust, runs in real time on commodity hardware, and is reproducible from a clean checkout.

We solved CartPole-v1 (avg=195.2 on the v0.5.0 codebase) using only local three-factor learning and developmental morphogenesis, with no backpropagation through the morphon network. On MNIST we reach \textbf{87.7\%} stateless accuracy in v4.6.0 (standard profile) using stateless training --- up from 30\% in v3.0.0 and from $\sim$50\% in v4.4.0 --- driven by three complementary advances: spike-timing inhibitory STDP (Vogels 2011) that breaks hub-morphon dominance (1 labeled class $\rightarrow$ 12); Pulse Kernel Lite that delivers a \textbf{4.3$\times$ wall-clock speedup} by fitting the fast-path arrays in L1 cache; and \emph{stateless training}, which resolves the co-adaptation problem identified in our v4.4.0 diagnostic analysis.

A key diagnostic finding of this work is the characterization of MORPHON as an LSM with co-adapted readout (Section~\ref{sec:lsm_characterization}). Offline classifiers on frozen morphon activations score 10.5\% (random chance), while the same activations yield 48.5\% online via the co-trained analog readout, and a sensory-only ablation confirms associative morphons contribute 35.5pp. This is not a failure of the reservoir --- it is the defining property of the LSM regime \cite{maass2002real}: features are discriminative only through the co-trained readout. The problem this creates for a per-image classifier is that sequential state carry-over contaminates the readout, co-adapting it to temporal patterns rather than per-image structure. Our sequential-vs-shuffled diagnostic directly confirmed this: the 18\% online accuracy of standard V3 rose to 44\% under stateless evaluation, with most of the remaining deficit attributable to temporal carry-over during test. Stateless training removes this confound at source, yielding 77\% on the fast profile and 87.7\% on the standard profile. The two-sentence technical lesson is: \emph{when training a per-image classifier on a stateful substrate, the training protocol must match the evaluation protocol; sequential training of a stateless classifier is a category error}.

We documented four failure modes encountered during development. Three (modulatory explosion, motor silencing, LTD vicious cycle) have well-known biological analogues and biology-derived fixes. The fourth --- premature ``Mature'' stage detection on dense-reward classification --- is, to our knowledge, novel and broadly applicable to any biologically-inspired homeostatic system that uses raw reward as a maturity signal. The fix is to use reward prediction error instead, which is the principled approach all such systems should adopt. A fifth failure mode (Section~\ref{sec:istdp_timescale}) is the \emph{iSTDP timescale mismatch}: computing inhibitory weight updates on a slow running average allows hub morphons to entrench before corrective inhibition can respond. A sixth failure mode, documented here for the first time, is \emph{novelty gain collapse}: Endoquilibrium's energy-pressure rule was suppressing the novelty modulation channel ($\mathit{ng} \rightarrow 0.32$), throttling three-factor STDP mid-training. The fix --- preserving $\mathit{ng}$ under energy pressure while suppressing only plasticity multiplication --- holds $\mathit{ng}$ at $\sim$1.60 throughout training and synergizes with stateless training to produce the v4.6.0 accuracy gains.

Post-training diagnostics further reveal that the v3.0.0 self-healing result --- a 30\% intact network recovering to 48\% after 30\% ablation --- was a diagnostic rather than a feature. Damage cleared entrenched hub morphons that rate-based iSTDP had failed to suppress; spike-timing iSTDP provides the equivalent plasticity regime from the start, absorbing the entire +18pp gain into the intact baseline. The general principle remains: \emph{the limiting factor in a developmental system is often a specific failure of the regulatory mechanism, not a fundamental capacity limit of the substrate}. Finding and fixing that failure is worth more than any amount of post-hoc damage recovery.

The MI engine, the experiments in this paper, and the full failure-mode catalogue are available at \texttt{https://github.com/SimplyLiz/Morphon}. This preprint is archived at Zenodo with DOI \texttt{10.5281/zenodo.19466544}.

\paragraph{Acknowledgments.}
The authors thank the Rust ecosystem, the petgraph crate, and the Diehl \& Cook MNIST baseline that calibrated our expectations.
